Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2025-01-14 paper/main.tex
minor cleanup to fig 6 across-lecture results

Finally, we estimated participants' knowledge for each question about one
lecture using their performance on questions (from the same quiz) about the
\textit{other} lecture. This is an especially stringent test of our approach.
Our primary assumption in building our knowledge estimates is that knowledge
about a given concept is similar to knowledge about other concepts that are
nearby in the embedding space. However, our analyses in Figure~\ref{fig:topics}
and Supplementary Figure~\topicWeights~show that the embeddings of content from
the two lectures are largely distinct. Therefore any predictive power of the
knowledge estimates must overcome large distances in the embedding space. To
put this in concrete terms, this test requires predicting participants'
performance on individual highly specific questions about the formation of
stars, using each participants' responses to just five multiple choice
questions about the fundamental forces of the universe (and vice versa).

We found that, before viewing either lecture (i.e., on Quiz~1), participants'
abilities to answer \textit{Four Fundamental Forces}-related questions could
not be predicted from their responses to \textit{Birth of Stars}-related
questions ($OR = 1.896,\ 95\%\ \textnormal{CI} = [0.419,\ 9.088],\ \lambda_{LR}
= 0.712,\ p = 0.404$), nor could their abilities to answer \textit{Birth of
Stars}-related questions be predicted from their responses to \textit{Four
Fundamental Forces}-related questions ($OR = 1.522,\ 95\%\ \textnormal{CI} =
[0.332,\ 6.835],\ \lambda_{LR} = 0.286,\ p = 0.611$). Similarly, we found that
participants' performance on questions about either lecture could not be
predicted given their responses to questions about the other lecture after
viewing \textit{Four Fundamental Forces} but before viewing \textit{Birth of
Stars} (i.e., on Quiz~2; \textit{Four Fundamental Forces} questions given
\textit{Birth of Stars} questions: $OR = 3.49,\ 95\%\ \textnormal{CI} =
[0.739,\ 12.849],\ \lambda_{LR} = 3.266,\ p = 0.083$; \textit{Birth of Stars}
questions given \textit{Four Fundamental Forces} questions: $OR = 2.199,\ 95\%\
\textnormal{CI} = [0.711,\ 5.623],\ \lambda_{LR} = 2.304,\ p = 0.141$). Only
after viewing \textit{both} lectures (i.e., on Quiz~3) did these across-lecture
knowledge estimates reliably predict participants' success on individual quiz
questions (\textit{Four Fundamental Forces} questions given \textit{Birth of
Stars} questions: $OR = 11.294,\ 95\%\ \textnormal{CI} = [1.375,\ 47.744],\
\lambda_{LR} = 10.396,\ p < 0.001$; \textit{Birth of Stars} questions given
\textit{Four Fundamental Forces} questions: $OR = 7.302,\ 95\%\ \textnormal{CI}
= [1.077,\ 44.879],\ \lambda_{LR} = 4.708,\ p = 0.038$). Taken together, our
results suggest that our ability to form estimates solely across different
content areas is more limited than our ability to form estimates that
incorporate responses to questions across both content areas (as in
Fig.~\ref{fig:predictions}, ``All questions'') or within a single content area (as
in Fig.~\ref{fig:predictions}, ``Within-lecture'').  However, if participants have recently
received some training on both content areas, the knowledge estimates appear to be informative
even across content areas.

We speculate that these ``Across-lecture'' results might relate to some of our
earlier work on the nature of semantic representations~\citep{MannKaha12}. In
that work, we asked whether semantic similarities could be captured through
behavioral measures, even if participants' ``true'' internal representations
differed from the embeddings used to \textit{characterize} participants'
behaviors. We found that mismatches between someone's internal representation
of a set of concepts and the representation used to characterize their
behaviors can lead to underestimates of how semantically driven their behaviors
are. Along similar lines, we suspect that in our current study, participants'
conceptual representations may initially differ from the representations
learned by our topic model. (Although the topic models are still
\textit{related} to participants' initial internal representations; otherwise
we would have found that knowledge estimates derived from Quiz 1 and 2
responses would have no predictive power in the other tests we conducted.)
After watching both lectures, however, participants' internal representations
may become more aligned with the embeddings used to estimate their knowledge
(since those embeddings were trained on the lecture transcripts). This could
help explain why the knowledge estimates derived from Quizzes 1 and 2 (before
both lectures had been watched) do not reliably predict performance across
content areas, whereas estiamtes derived from Quiz 3 \textit{do} reliably
predict performance across content areas.

>>>>

Finally, we estimated participants' knowledge for each question about one
lecture using their performance on questions (from the same quiz) about the
\textit{other} lecture. This is an especially stringent test of our approach.
Our primary assumption in building our knowledge estimates is that knowledge
about a given concept is similar to knowledge about other concepts that are
nearby in the embedding space. However, our analyses in Figure~\ref{fig:topics}
and Supplementary Figure~\topicWeights\ show that the embeddings of content from
the two lectures (and of their associated quiz questions) are largely distinct 
from each other. Therefore, any predictive power of these across-lecture 
knowledge estimates must overcome large distances in the embedding space. To put 
this in concrete terms, this test requires predicting participants' performance 
on individual, highly specific questions about the formation of stars from their 
responses to just five multiple-choice questions about the fundamental forces of 
the universe (and vice versa).

We found that, before viewing either lecture (i.e., on Quiz~1), participants'
abilities to answer \textit{Four Fundamental Forces}-related questions could
not be predicted from their responses to \textit{Birth of Stars}-related
questions ($OR = 1.896,\ 95\%\ \textnormal{CI} = [0.419,\ 9.088],\ \lambda_{LR}
= 0.712,\ p = 0.404$), nor could their abilities to answer \textit{Birth of
Stars}-related questions be predicted from their responses to \textit{Four
Fundamental Forces}-related questions ($OR = 1.522,\ 95\%\ \textnormal{CI} =
[0.332,\ 6.835],\ \lambda_{LR} = 0.286,\ p = 0.611$). Similarly, we found that
participants' performance on questions about either lecture could not be
predicted given their responses to questions about the other lecture after
viewing \textit{Four Fundamental Forces} but before viewing \textit{Birth of
Stars} (i.e., on Quiz~2; \textit{Four Fundamental Forces} questions given
\textit{Birth of Stars} questions: $OR = 3.49,\ 95\%\ \textnormal{CI} =
[0.739,\ 12.849],\ \lambda_{LR} = 3.266,\ p = 0.083$; \textit{Birth of Stars}
questions given \textit{Four Fundamental Forces} questions: $OR = 2.199,\ 95\%\
\textnormal{CI} = [0.711,\ 5.623],\ \lambda_{LR} = 2.304,\ p = 0.141$). Only
after viewing \textit{both} lectures (i.e., on Quiz~3) did these across-lecture
knowledge estimates reliably predict participants' success on individual quiz
questions (\textit{Four Fundamental Forces} questions given \textit{Birth of
Stars} questions: $OR = 11.294,\ 95\%\ \textnormal{CI} = [1.375,\ 47.744],\
\lambda_{LR} = 10.396,\ p < 0.001$; \textit{Birth of Stars} questions given
\textit{Four Fundamental Forces} questions: $OR = 7.302,\ 95\%\ \textnormal{CI}
= [1.077,\ 44.879],\ \lambda_{LR} = 4.708,\ p = 0.038$). Taken together, these
results suggest that our ability to form estimates solely across different
content areas is more limited than our ability to form estimates that
incorporate responses to questions from both content areas (as in 
Fig.~\ref{fig:predictions}, ``All questions'') or within a single content area 
(as in Fig.~\ref{fig:predictions}, ``Within-lecture''). However, if participants 
have recently received some training on both content areas, the knowledge 
estimates appear to be informative even across content areas.

We speculate that these ``Across-lecture'' results might relate to some of our
earlier work on the nature of semantic representations~\citep{MannKaha12}. In 
that work, we asked whether semantic similarities could be captured through 
behavioral measures, even if participants' ``true'' internal representations 
differed from the embeddings used to \textit{characterize} their behaviors. We 
found that mismatches between an individual's internal representation of a set 
of concepts and the representation used to characterize their behaviors can lead 
to underestimates of how semantically driven those behaviors are. Along similar 
lines, we suspect that in our current study, participants' conceptual 
representations may initially differ from the representations learned by our 
topic model. (Although the topic model's representations are still 
\textit{related} to participants' initial internal representations; otherwise we 
would have found that knowledge estimates derived from Quizzes~1 and 2 had no 
predictive power in the other tests we conducted.) After watching both lectures, 
however, participants' internal representations may become more aligned with the 
embeddings used to estimate their knowledge (since those embeddings were trained 
on the lectures' transcripts). This could help explain why the knowledge 
estimates derived from Quizzes~1 and 2 (before both lectures had been watched) 
do not reliably predict performance across content areas, whereas estimates 
derived from Quiz~3 do.
